\subsection{Mutex Model}
\label{mutex-model}

{\it Mutexes} are a mechanism for
achieving mutual exclusion among a set of concurrently executing threads.  
A mutex is an object that can be in one of two states, {\it locked} and 
{\it unlocked}.  Mutex objects must support {\it lock} and {\it unlock},
which move the mutex from the unlocked state to the locked state, and
vice versa. Mutual exclusion is guaranteed since the lock and unlock
operations are atomic and only one contender will successfully acquire
the lock. Attempts to lock a mutex that is already locked by some 
other thread will cause the calling thread to wait until the mutex is 
unlocked.  

\paragraph{Fluke {\tt mutex} Objects}

The Fluke implementation of mutexes uses Fluke wait queues to keep track
of threads waiting to acquire a particular mutex.

\paragraph{The Promela {\tt Mutex} Type}

Since Promela has builtin atomicity support via the {\tt atomic}
and {\tt d_step} constructs, it was not necessary for us to build our
model Mutexes on top of our model WaitQueues. 
Rather, one bit is sufficient to represent the state of a mutex.
We call Mutexes with one state bit {\it simple} mutexes. They are
implemented in {\tt src/fluke/Mutex-simple.pr}.

However, there are many useful properties that one may wish to verify
about mutexes and how they are used by other code.  For example, it may
be desirable to assert that the same thread that locked a mutex must unlock
it. Also, we might wish to
ensure that the {\it same} thread does not try to lock a mutex it already
holds since this action would result in deadlock.  
We have a second, more heavyweight,
implementation of Promela Mutexes, called {\it safe} mutexes that address
these concerns and aggressively make sanity assertions.  This implementation
can be found in {\tt src/fluke/Mutex-safe.pr}.

Both implementations of Promela Mutexes support the same set of functions.
The intention is that {\tt Mutex-safe.pr} will be used to verify the
appropriate use of Mutexes in a system, and then {\tt Mutex-simple.pr}
will be substituted when the whole system is tested, to minimize state
vector length.
