\section{Fluke Core Models}
\label{core}

While evolving our current Promela model of Fluke
IPC we have developed a support infrastructure which includes Promela
implementations of many low-level Fluke objects used during Fluke IPC 
sessions. We call this support infrastructure the
{\it Core Models}.  These models model such Fluke entities as mutexes,
condition variables, wait queues, links, ports and port sets, and the
Fluke kernel entry layer (which embodies behavior intimately related to
the cancellability of IPC operations).

Implementation details and the design rationale of these models are
presented here.  Where appropriate, the implementation of these models is
quite faithful to the C-language source from which they are derived.  When
particular divergence from the C-code has occurred, we note the underlying
motivation and discuss the semantic similarities between the two
implementations.

Since our model is heavily based on the actual implementation, we have
a high measure of confidence that the results obtained from model
checking are in fact applicable to the real system.

\subsection{Translation Artifacts and the Sliding {\tt d_step} Technique}

Since Promela does not support language-level procedures or named blocks,
we were forced to rely on C preprocessor macros to give us the ability to
name code segments and provide a function-like decomposition of source
code.  Since these macros are not lexical scopes, and many instances of
the same macro can be instantiated in a single lexical scope (proctype),
these macros cannot contain variable declarations, labels or other symbol
declarations or duplicate symbol declaration will occur.  Thus, most macros
assume the presence of a local variable called {\tt rc} which
is used to hold the ``return value'' from the ``function'' that the macro
embodies.  This variable can be interrogated by the user of the macro.

The Core Models are designed to allow the use of a
technique we call the {\it sliding {\tt d_step}}.  When working with
a model as large as the Fluke IPC model, the problem of state-space
explosion is an omnipresent threat.  Since the Core Models are largely
self-contained and contain no external references, individual Core
Models (or groups of them) can be verified together, and then 
their ``public methods'' can be encased with the SPIN/Promela 
{\tt d_step} statement, which will avoid an overall increase in 
state-space. 
%% "can be kept at the edge of manageability" is too wordy... 
%% but I'll leave it in since I can't immediately come up with something better
The positioning of these {\tt d_steps} is quite flexible due to the
hierarchical and rigidly encapsulated structure of our models.  Thus,
problems can be kept at the edge of manageability, to verify the Fluke
implementation as rigorously as possible using the hardware available.

\input{core-mutex.tex}

\input{core-wq.tex}

\input{core-cond.tex}

\input{core-links.tex}

\input{core-portsEtc.tex}

\input{core-kentry.tex}
